---
title: "Econ 104 Course Project"
author: "Valeria Meza, Brianna Pineda, Sara Hameed"
date: "2026-02-11"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

library(tidyverse)
library(Boruta)
library(car)
library(lmtest)
library(sandwich)
library(tseries)
library(boot)
library(caret)

set.seed(123)
```

```{r load-data}
df <- read_csv("data/msa_housing_final.csv")
nrow(df)
glimpse(df)
```

```{r create-employment-rate}
# Create employment_rate if missing (so you have 10 predictors)
if (!"employment_rate" %in% names(df)) {
  df <- df %>% mutate(employment_rate = 1 - unemployment_rate)
}

"employment_rate" %in% names(df)
```


```{r define-vars}
y_var <- "med_home_value"

quant_vars <- c(
  "med_income",
  "log_income",
  "population",
  "log_population",
  "unemployment_rate",
  "employment_rate",
  "med_age",
  "age_squared",
  "vacancy_rate",
  "college_share"
)

cat("Missing variables:", setdiff(c(y_var, quant_vars), names(df)), "\n")
```
## Introduction
  Housing markets are an essential indicator of economic development and stability as they reflect a region’s health and development. Therefore, the impact of local socio-economic factors and demographic characteristics on housing prices is integral to understanding the development of local economies. Metropolitan statistical areas (MSAs) serve as a useful framework for housing markets as they integrate suburban outer regions and urban centers. As defined by the U.S. Census Bureau, they are a combination of multiple counties linked by population patterns and commuting routes providing a holistic view of population, demographic, and economic factors that may impact housing prices(2023). This paper examines data from the American Community Survey conducted by the U.S. Census Bureau which collects national data on various parameters that detail social, economic, and demographic characteristics across the country every year. It spans many different topics including education, employment, and housing allowing for the analysis of the intersection between these factors. 

## 1. Economic Theory and Hypothesis
### (a) Research Question
  How are housing prices in metropolitan statistical areas (MSAs) in the United States affected by local economic and demographic factors such as unemployment rate, vacancy rates,and  median household income?
### (b) Economic Theory
     Economic theory suggests that constrained metropolitan housing supply causes higher housing prices. The elasticity of housing supply plays a key role in determining the price of housing. In a traditional supply and demand model, supply adjusts to demand changes based on its elasticity. With elastic supply, production races to meet increased demand. In metropolitan housing markets, however, housing supply is rather inelastic. As city populations grow, demand for housing follows suit. When the demand for more housing in big cities increases, barriers such as zoning laws, environmental regulations, and a lack of available land make it hard to expand. In the long run, acquiring permits, financing, and construction are often multi-year endeavors, adding to the inability of housing supply to quickly meet new demand. Supply of housing is inelastic and unable to change very much in response to demand shocks like migration and income growth. In condensed cities, housing supply is especially inelastic because of limited space for construction (Saiz, 2010). Rather than increasing the number of homes through construction, house values inflate to reach equilibrium. Therefore, metropolitan areas with larger populations tend to exhibit higher home values.   
     Housing prices depend strongly on local labor market conditions. Shocks to the labor market can act as a strong influence on the demand for housing in urban areas. When there’s a negative shock to the job market, in the form of rising unemployment, we can expect a negative effect on housing prices. The demand for housing is dependent on consumer income, and when there’s a negative shock to the labor market, reducing employment and income, the housing demand curve shifts inward. Unemployment leads to fewer qualified mortgage applicants, effectively driving the buyer market for housing down. Less job stability and more income insecurity bode poorly for borrowers seeking to secure loans to purchase property (Mian and Sufi, 2011).  Further, when unemployed people can’t find work in the city, they may be tempted to migrate toward more fruitful labor markets, reducing local housing demand (Blanchard and Katz, 1992). When purchasing power and growth opportunities are limited, property purchases plummet, driving down prices. Therefore, metropolitan areas with higher unemployment rates yield lower median house prices. 
     In the long run, persistent differences in workforce education levels play a role in creating differences in housing demand across metropolitan regions. Economic theory suggests that high-skill workers or those with higher education attainment gravitate toward productive metropolises with higher wages and more employment opportunities, even when housing is more expensive (Diamond, 2016). This sorting pattern creates high-skill clusters in cities with firms that require such workers. Those within these clusters earn higher income and increased purchasing power, so their willingness to pay for more expensive homes rises. With a limited housing supply and high demand for property, competitor buyers bid up the values of homes. As a result, these metropolises with clusters of highly educated workers sustain higher median housing prices.

### (c) Hypotheses
  Metropolitan statistical areas (MSAs) with higher socio-economic status–characterized by lower unemployment rate, higher median incomes, and higher education attainment–demonstrate higher median housing prices. 
### (d) Expected Signs
  The coefficient of unemployment rate is expected to be negative because greater unemployment reduces household income as well as the ability to borrow leading to less demand for housing. With diminished ability to pay for housing, either due, housing prices reflect lower purchasing power. Similarly, as vacancy rates increase, housing prices decrease due to greater supply of housing units which would lead to a negative coefficient to highlight its negative relationship with housing values. 
  Higher median incomes are associated with higher housing prices due to the income effect. Higher salaries lead to greater purchasing power for consumers as well as improved access to credit which leads to greater demand for higher priced housing. Thus, the coefficient for median income would be positive to reflect the positive effect. In like manner, higher college rates are associated with the potential for larger incomes resulting in a positive coefficient for college rates. Population would also exhibit a positive effect on housing incomes as it would place constraints on housing supply, leading to higher median housing prices. 
  Even though factors such as land-use restrictions could constrain housing supply, these are the expected effects of the most influential variables from the data set. 



## 2. Variable Selection 
### (a) Boruta Algorithm (Quantitative Predictors)  
```{r boruta}
# Run Boruta algorithm 
boruta_df <- df %>%   
  dplyr::select(dplyr::all_of(c(y_var, quant_vars))) %>%
  tidyr::drop_na()  

set.seed(123) 
Bor.res <- Boruta(   
  x = boruta_df[, quant_vars],   
  y = boruta_df[[y_var]],   
  doTrace = 0,   
  maxRuns = 100 )  

# Fix tentative variables 
Bor.res.fixed <- TentativeRoughFix(Bor.res)  
# Selected variables 
selected_vars <- getSelectedAttributes(Bor.res.fixed, withTentative = FALSE) 
selected_vars  

# Plot results
plot(Bor.res.fixed, xlab = "", xaxt = "n",      
     main = "Boruta Algorithm: Feature Importance") 
lz <- lapply(1:ncol(Bor.res.fixed$ImpHistory), function(i)   Bor.res.fixed$ImpHistory[is.finite(Bor.res.fixed$ImpHistory[, i]), i])

names(lz) <- colnames(Bor.res.fixed$ImpHistory) 
Labels <- sort(sapply(lz, median))  

axis(side = 1, las = 2, labels = names(Labels),      
     at = 1:ncol(Bor.res.fixed$ImpHistory), cex.axis = 0.8)
```
We applied the Boruta algorithm to identify the most relevant quantitative predictors of median home value across U.S. metropolitan statistical areas. The algorithm compares the importance of each predictor to randomized shadow variables and retains only those that consistently outperform the shadows. Using 100 iterations and a fixed random seed for reproducibility, Boruta confirmed all ten candidate quantitative predictors as important. These include measures of income, population size, labor market conditions, demographic structure, housing market tightness, and educational attainment.

### (b) Factor/indicator variables
```{r create-indicators}

# Create Indicator Variables
# 1. High-income MSA (above median income)
income_cutoff <- median(df$med_income, na.rm = TRUE)

df <- df %>%
  mutate(
    high_income_msa = ifelse(med_income > income_cutoff, 1, 0)
  )

table(df$high_income_msa)


# 2. Large MSA (above median population)
population_cutoff <- median(df$population, na.rm = TRUE)

df <- df %>%
  mutate(
    large_msa = ifelse(population > population_cutoff, 1, 0)
  )

table(df$large_msa)


# 3. Tight housing market (below median vacancy rate)
vacancy_cutoff <- median(df$vacancy_rate, na.rm = TRUE)

df <- df %>%
  mutate(
    tight_market = ifelse(vacancy_rate < vacancy_cutoff, 1, 0)
  )

table(df$tight_market)
```

```{r define-factor-vars}

factor_vars <- c(
  "high_income_msa",
  "large_msa",
  "tight_market"
)

factor_vars
```
The dataset did not contain existing categorical predictors beyond the MSA name. Therefore, we constructed economically meaningful binary indicators. We created a high-income MSA indicator (above the median income), a large metropolitan area indicator (above median population), and a tight housing market indicator (below median vacancy rate). These variables allow us to test whether structural differences across metropolitan areas are associated with variation in housing prices. 

### (c) Final predictor set
```{r final-predictor-set}

final_predictors <- c(quant_vars, factor_vars)

final_predictors
```

## 3. Descriptive Analysis and EDA
###(a) Provide detailed description of your dataset: source, time period, sample selection, vari- able definitions 
We are analyzing the cross-sectional variation in housing prices across various U.S. Metropolitan Statistical Areas (MSAs) using the American Community Survey (ACS) 5-Year Estimates for 2019-2023. The unit we are observing is the MSA and the final sample includes 393 MSAs. There are no missing values for the outcome variable. The dependent variable is the median home value (med_home_value). We are using the explanatory variables to capture local economic conditions, market size, housing market tightness, and demographic/human capital composition. A few variables are transformed to explain non-linearity and scale differences. 

Variable Definitions on MSA level: 
-med_home_value: median value of owner-occupied housing units (outcome) 
-med_income: median household income 
-population: total population 
-unemployment_rate: unemployed / labor_force 
-employment_rate: 1 − unemployment_rate (constructed) 
-vacancy_rate: vacant_units / housing_units 
-college_share: college_grad / college_total 
-med_age: median age 
-log_income: ln(med_income)
-log_population: ln(population) 
-age_squared: med_age² 

###(b) Begin by providing a descriptive analysis of your variables. This should include things like histograms, quantile plots, correlation plots, boxplots, scatterplots, etc.
```{r eda-vars}
# Outcome + predictors we decided to keep
eda_vars <- c(
  y_var,
  "med_income","log_income",
  "population","log_population",
  "unemployment_rate","employment_rate",
  "med_age","age_squared",
  "vacancy_rate","college_share"
)

# Keep only numeric columns (should be all of them here)
eda_vars <- eda_vars[eda_vars %in% names(df)]
eda_vars
```
###STEP 1 — Histograms (univariate)
```{r eda-histograms}
# Histograms of outcome and quantitative predictors 
eda_df <- df %>%   
  dplyr::select(dplyr::all_of(eda_vars)) %>%   
  tidyr::drop_na()  

par(mfrow = c(2, 3))   # adjust depending on number of variables 
for (v in eda_vars) {   
  truehist(eda_df[[v]],            
           xlab = v,            
           ylab = "Density",            
           main = paste("Histogram of", v),            
           col = "skyblue3") 
  } 
par(mfrow = c(1, 1))
```

###STEP 2 — Quantile plots (QQ plots)
```{r eda-qqplots} 
eda_df <- df %>%   
  dplyr::select(dplyr::all_of(eda_vars)) %>%   
  tidyr::drop_na()  

par(mfrow = c(2,3))  # adjust depending on number of variables  
for(v in eda_vars){    
  qqnorm(eda_df[[v]],          
         col = "skyblue4",          
         pch = 20,          
         lwd = 1,          
         main = paste("QQ Plot:", v))    
  qqline(eda_df[[v]],          
         col = "red",          
         lwd = 2)  }  
par(mfrow = c(1,1))
```
###STEP 3 — Correlation plot (multivariate)
```{r eda-correlation}
num_df <- df %>% dplyr::select(all_of(eda_vars)) %>% drop_na()

cor_mat <- cor(num_df)
round(cor_mat, 3)
```
```{r eda-corr-plot}
GGally::ggcorr(num_df, label = TRUE, label_round = 2) +
  labs(title = "Correlation Plot (Numeric Variables)")
```
### STEP 4 — Boxplots
```{r eda-boxplots}
box_df <- df %>% 
dplyr::select(dplyr::all_of(eda_vars)) %>%
tidyr::drop_na()

par(mfrow = c(3,1))  # adjust if you have more variables

for(v in eda_vars){

boxplot(box_df[[v]],
main = paste("Boxplot of", v),
col = "skyblue3",
ylab = v)

}

par(mfrow = c(1,1))
```
### STEP 5 — Scatterplots vs the outcome (med_home_value)
```{r eda-scatter-outcome}
predictors <- setdiff(eda_vars, y_var)

for (v in predictors) {
  p <- ggplot(df, aes(x = .data[[v]], y = .data[[y_var]])) +
    geom_point(alpha = 0.5) +
    geom_smooth(method = "lm", se = FALSE) +
    labs(title = paste("Scatter:", y_var, "vs", v),
         x = v, y = y_var)
  print(p)
}
```
-Histograms depict the distribution shape and show if logs help. 
-QQ Plots check for normality and flag extreme tails and outliers.
-Correlation Plots highlight multicollinearity. 
-Boxplots show outliers and variables that are dispersed in an unusual way. 
-Scatterplots vs. med_home_value shows the direction and possible nonlinear patterns that motivate transformations. 

###(c) Create a comprehensive descriptive statistics table (mean, median, std dev, min, max) 
```{r eda-descriptive-table}
# Variables to summarize
eda_vars <- c(
"med_home_value",
"med_income", "log_income",
"population", "log_population",
"unemployment_rate", "employment_rate",
"med_age", "age_squared",
"vacancy_rate",
"college_share"
)

# Keep only selected variables
desc_df <- df %>%
dplyr::select(dplyr::all_of(eda_vars))

# Create descriptive statistics table
desc_table <- data.frame(
Variable = names(desc_df),
Mean = sapply(desc_df, mean, na.rm = TRUE),
Median = sapply(desc_df, median, na.rm = TRUE),
SD = sapply(desc_df, sd, na.rm = TRUE),
Min = sapply(desc_df, min, na.rm = TRUE),
Max = sapply(desc_df, max, na.rm = TRUE),
N = sapply(desc_df, function(x) sum(!is.na(x)))
)

# Round results
desc_table[, -1] <- round(desc_table[, -1], 3)

# Print table
print(desc_table)
```

###(d) Estimate density plots for all your variables, and show the respective fitted distributions 
### 1) Density plots for all variables
```{r eda-density}
# Density plots for all variables
dens_df <- df %>%
dplyr::select(dplyr::all_of(eda_vars)) %>%
tidyr::drop_na()

par(mfrow = c(3, 4))  # adjust depending on number of variables

for (v in eda_vars) {
plot(density(dens_df[[v]]),
main = paste("Density Plot of", v),
xlab = v,
lwd = 2,
col = "skyblue4")
}
par(mfrow = c(1, 1))
```
### 2) Fit disttributions 
```{r eda-fit-distributions}
# Density plots with fitted normal distributions
fit_df <- df %>%
dplyr::select(dplyr::all_of(eda_vars)) %>%
tidyr::drop_na()

par(mfrow = c(3, 4)) # adjust depending on number of variables

for (v in eda_vars) {

x <- fit_df[[v]]

# Plot actual density
plot(density(x),
main = paste("Density Fit:", v),
xlab = v,
lwd = 2,
col = "red")

# Add fitted normal density
curve(dnorm(x, mean = mean(x), sd = sd(x)),
add = TRUE,
col = "black",
lwd = 2)

legend("topright",
legend = c("Actual", "Normal"),
col = c("red", "black"),
lwd = 2,
bty = "n")
}

par(mfrow = c(1, 1))
```
The density plot shows that several variables, especially median home value and population, are right-skewed. The fitted normal distributions do not fully capture these shapes. This indicates the departure from normality on some variables. 

###(e) Identify if there are any non-linearities within your variables. What transformations should you perform to make them linear? What would happen if you included nonlinear variables in your regression models without transforming them first? 
### 1) Scatterplots with LOESS
```{r eda-nonlinearity-scatter}
predictors_to_check <- c(
  "med_income", "population",
  "unemployment_rate", "vacancy_rate",
  "med_age", "college_share"
)

df_scatter <- df %>%
  dplyr::select(med_home_value, dplyr::all_of(predictors_to_check)) %>%
  drop_na()

df_long_scatter <- df_scatter %>%
  pivot_longer(cols = -med_home_value, names_to = "xvar", values_to = "x")

ggplot(df_long_scatter, aes(x = x, y = med_home_value)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~xvar, scales = "free", ncol = 2) +
  labs(title = "Nonlinearity check: Outcome vs predictors")
```
Many of the relationships depicted are nonlinear, especially for population and income, due to the right-skewness and diminishing marginal effects. That is why we use log transformations for population and income and allow curvature with age_squared. If nonlinear variables are included in regression models without transfromations, they model can be misspecified and could fail to capture the true relationship between the variables, leading to biased coefficient estimates. 

###(f) Comment on any outliers and/or unusual features of your variables, and then justify their removal, exclusion, or imputation
### 1) Outlier counts by variable (IQR rule) 
```{r eda-outliers-iqr} 
# Function to count IQR outliers 
iqr_outlier_counts <- function(x){    
  q1 <- quantile(x, 0.25, na.rm = TRUE)   
  q3 <- quantile(x, 0.75, na.rm = TRUE)    
  
  iqr <- q3 - q1    
  lower <- q1 - 1.5*iqr   
  upper <- q3 + 1.5*iqr    
  
  sum(x < lower | x > upper, na.rm = TRUE) }  

# Create vector to store counts 
outlier_counts <- numeric(length(eda_vars))  

# Loop through variables 
for(i in 1:length(eda_vars)){   
  outlier_counts[i] <- iqr_outlier_counts(df[[eda_vars[i]]]) }  

# Create results table 
outlier_summary <- data.frame(   
  Variable = eda_vars,   
  Outlier_Count = outlier_counts )  

# Sort from most to least outliers 
outlier_summary <- outlier_summary[order(-outlier_summary$Outlier_Count),]
print(outlier_summary)
```
### 2) Biggest outliers for population + med_home_value
```{r eda-outliers-top} 
# Look at largest observations for key variables 
outlier_check <- df[, c("NAME","med_home_value","population","vacancy_rate","med_income")]

# Sort by population (largest first) 
outlier_check <- outlier_check[order(-outlier_check$population), ]

# Show top 10 
head(outlier_check, 10)
```
Population and home vales showcase extreme observations that correspond to very large metro areas like L.A. and NYC. They are not data errors, they are real-world variation, so we retain the values. To reduce their impact and improve linearity, we use log transformations (log_population and log_income) and use robust standard errors later on. 

###(g) If you have any NAs, remove them or impute them using any of the methods discussed in class, but make sure to justify your choice
### 1) Missingness table 
```{r eda-missingness}
na_table <- data.frame(
  Variable = names(df),
  NA_Count = sapply(df, function(x) sum(is.na(x))),
  NA_Share = sapply(df, function(x) mean(is.na(x)))
)

na_table$NA_Share <- round(na_table$NA_Share, 3)

na_table <- na_table[order(-na_table$NA_Count), ]

print(na_table)
```
There are no missing values detected in the data set. Therefore, there is no deletion of values needed and all observations were retained for analysis. 

## 4. Model Building and Selection

### 1) Identify baseline model

### We estimate the following baseline model: 

$$
\text{med home value}_i =
\beta_0
+ \beta_1 \log(\text{income}_i)
+ \beta_2 \log(\text{population}_i)
+ \beta_3 (\text{unemployment rate}_i)
+ \beta_4 (\text{vacancy rate}_i)
+ \beta_5 (\text{college share}_i)
+ \beta_6 (\text{med age}_i)
+ \beta_7 (\text{age squared}_i)
+ u_i
$$

```{r base regression}

base_model <- lm(med_home_value ~ log_income + log_population + unemployment_rate + vacancy_rate + college_share + med_age + age_squared, data = df)
summary(base_model)
```
### 2) Transform Dependent Variable

```{r med_home_value histogram}
hist(df$med_home_value)

```
The distribution is heavily skewed right, warranting a log transformation. Median house values are positive and will not interfere with a log transformation.  

We'll now estimate the following log-log model:

$$
\log(\text{med home value}_i) =
\beta_0
+ \beta_1 \log(\text{income}_i)
+ \beta_2 \log(\text{population}_i)
+ \beta_3 (\text{unemployment rate}_i)
+ \beta_4 (\text{vacancy rate}_i)
+ \beta_5 (\text{college share}_i)
+ \beta_6 (\text{med age}_i)
+ \beta_7 (\text{age squared}_i)
+ u_i
$$
```{r log model}
Y <- log(df$med_home_value)
model2 <- lm(Y ~ log_income + log_population + unemployment_rate + vacancy_rate + college_share + med_age + age_squared, data = df)
summary(model2)
```
### 3) Test for multicollinearity using Variance Inflation Factors (VIF)
 
```{r VIF}
vif_values <- vif(base_model)
vif_values
```
### Plot VIF values

```{r VIF plot}
barplot(vif_values, main="VIF Values", col='skyblue3', ylim=c(0, 12))
abline(h=4.0, lwd=3, lty=2, col='red4')
abline(h=10, lwd=3, lty=2, col='gold4')
text(0.7, 5, "VIF 4-10", col='red4', pos=3)
text(0.7, 11, "VIF > 10", col='gold4', pos=3)
```
Median age and age squared, as expected, have high VIF values > 50. To mitigate this, we will center age about the mean.

```{r center age squared}
med_age_c <- df$med_age - mean(df$med_age)
age_c_2 <- med_age_c^2
```

### Estimate a new linear-log model with centered age squared
```{r model3.1}
model3.1 <- lm(med_home_value ~ log_income + log_population + unemployment_rate + vacancy_rate + college_share + med_age + age_c_2, data = df)
summary(model3.1)
```
### Estimate a new log-log model with centered age squared
```{r model 3.2}
model3.2 <- lm(Y ~ log_income + log_population + unemployment_rate + vacancy_rate + college_share + med_age + age_c_2, data = df)
summary(model3.2)
vif_values2 <- vif(model3.2)
print(vif_values2)
```
```{r VIF Plot 2}
barplot(vif_values2, main="VIF Values", col='skyblue3', ylim=c(0, 12))
abline(h=4.0, lwd=3, lty=2, col='red4')
abline(h=10, lwd=3, lty=2, col='gold4')
text(0.7, 5, "VIF 4-10", col='red4', pos=3)
text(0.7, 11, "VIF > 10", col='gold4', pos=3)
```
Multicollinearity between median age and age squared has been mitigated by centering age squared around its mean. 

### 4) Perform RESET Test to test for misspecification

```{r RESET Test}
resettest(base_model)
resettest(log_model)

```
Both models fail RESET Test, signifying misspecification. This may call for an interaction term between income and education level. Based on economic theory, individuals with higher education levels earn higher incomes, influencing demand for high-value homes.

$$
\log(\text{med home value}_i) =
\beta_0
+ \beta_1 \log({income}_i)
+ \beta_2 \text{college share}_i
+ \beta_3 \big[\log(\text{income}_i)\times \text{college share}_i\big]
+ \beta_4 \log(\text{population}_i)
+ \beta_5 \text{unemployment rate}_i
+ \beta_6 \text{vacancy rate}_i
+ \beta_7 \text{age}_i
+ \beta_8 \text{age sq}_i
+ u_i
$$

### Interaction Model


```{r Interaction Model}
interaction_model <- lm(Y ~ log_income*college_share +log_population + unemployment_rate + vacancy_rate + med_age_c + age_c_2,
  data = df)
summary(interaction_model)
resettest(interaction_model)
```
The interaction model with log(income) times college share still fails RESET test. However, given the large cross-sectional data, over-specifying may not improve AIC and BIC implications.

```{r AIC & BIC}
AIC(base_model, model3.2, interaction_model)
BIC(base_model, model3.2, interaction_model)
```
Based on AIC and BIC, Model3.2 is strictly preferred. Adding an interaction between income and college share only improved AIC by a value of 1.86 and BIC by a value of 5.84, reducing need for over-specification. 

### 5) Diagnostic Plots

### Cook's Distance Plot

```{r Cook's Distance Plot}

cooks_d <- cooks.distance(model3.2)

n <- length(cooks_d)
threshold <- 4/n

plot(cooks_d,
    type = "h",
    main = "Cook's Distance",
    ylab = "Cook's Distance",
    xlab = "Observation Index")

abline(h=threshold, col="blue", lty=2)
```
In general, most of the MSAs lie below the threshold line which indicates stability. The presence of a few influential points above the threshold suggests that certain MSAs may skew the coefficient estimates of our model due to their size which can lead to high median housing prices and high incomes. For that reason, keeping them in our model despite their influence is important because it showcases the variation present. 

### Residuals vs. Fitted Plot
```{r Residuals vs. Fitted Plot}
plot(model3.2$fitted.values, model3.2$residuals,
     xlab = "Fitted Values",
     ylab = "Residuals",
     main = "Residuals vs. Fitted",
     pch = 19, 
     col = "cadetblue",
     cex = 0.6)

abline(h=0, col = "blue", lty = 2)
```

The Residuals vs. Fitted Plot demonstrates that there is no pattern to indicate heteroskedasticity. Although there is minor spread towards the right of the cluster, most of the observations cluster around zero with a few outliers. This satisfies the linerarity and homoskedasticity assumptions. 

### QQ-Plot
```{r QQ-Plot}
res <- residuals(model3.2)

qqnorm(res,
      main = "QQ Plot of Residuals",
      pch = 19, col = "cadetblue")

qqline(res, col = "blue", lwd = 2)

library(ggplot2)
ggplot(data = data.frame(residuals = residuals(model3.2)), aes(sample = residuals)) + 
      stat_qq(color = "cadetblue", size = 1.5) + 
      stat_qq_line(color = "blue", lwd = 1.5) +
      labs(title = "QQ Plot of Residuals") +
      theme_minimal() 
```
The QQ-Plot model stays close to the reference line, suggesting that residuals are approximately normal. However, at the tails, the  residuals start to stray from the line, showcasing that there are outliers that may influence coefficients. 

### Scale-Location Plot
```{r Scale-Location Plot}
plot(fitted(model3.2),
     sqrt(abs(rstandard(model3.2))), 
     xlab = "Fitted Values",
     ylab = "Sqrt(|Standardized residuals|)")

abline(h = 0, lty = 2)
```

The random scatter of the residuals shows that this model is homoskedastic. 

## 5. Diagnostics and Robustness
```{r diag-setup}
library(lmtest)
library(sandwich)
library(car)
library(ggplot2)
library(dplyr)

final_model <- model3.2  

resid_final  <- residuals(final_model) 
fitted_final <- fitted(final_model)
```

## 5.1 Residuals vs Fitted (linearity + heteroskedasticity patterns)
```{r diag-resid-fitted}
plot(final_model, which = 1)
```
This plot shows that residuals are centered around zero and there is no strong funnel shape. This suggests that the homoskedasticity assumption is not satosfied. There is a slight curvature in the LOESS line that indicates minor nonlinearity. 

## 5.2 Homoskedasticity
#(a) BP Test
```{r diag-bptest} 
bptest(final_model)
```
The Breusch-Pagan test rejects homoskedasticity (p < 0.05). This indicates heteroskedasticity. We will use HC standard erros for inference.  

#(b) Scale-Location Plot
```{r diag-scale-location} 
plot(final_model, which = 3)
``` 
The Scale-Location plot shows a mostly stable spread across the fitted values. However, it does show a slight curvature that suggests mild heteroskedasticity. 

### 5.3 Normality of residuals
# a) Q-Q plots
```{r diag-qq} 
plot(final_model, which = 2)
```
The Q-Q plot shows that residuals are approximately normal. It shows minor deviations in the tails. 

# b) Histogram of residuals
```{r diag-resid-hist} 
hist(resid_final, breaks = 30, main = "Histogram of Residuals", xlab = "Residuals")
```
The residual histogram is bell-shaped. 

# c) Jarque-Bera test
```{r diag-jb, message=FALSE} 
library(tseries) 
jarque.bera.test(resid_final)
```
The Jarque-Bera test rejects normality (p < 0.05). This suggests non-normal residuals. OLS remains unbiased and consistent with a large sample, but we rely on robust SEs for inference. 

### 5.4 Outliers and influence 
# a) Cook’s distance plot
```{r diag-cooks} 
plot(final_model, which = 4)
```
```{r cooks-threshold} 
n <- nobs(final_model) 
cooks_cut <- 4/n 
cooks_cut
```
```{r cooks-which} 
which(cooks.distance(final_model) > cooks_cut)
```
A few observations exceed the Cook's distance threshold, indicating potentially influential MSAs. These are expected in cross-sectional housing data since the values reflect real economic variation rather than mistakes. 

# b) Leverage analysis
```{r diag-leverage} 
plot(final_model, which = 5)
```
The residuals vs. leverage plot shows that most observations have low leverage and residuals centered near zero, showcasing that the model fits the majority of MSAs well. A few observations, such as 383, show high leverage which means these observations represent metropolitan areas with extreme characteristics. 

### 5.5 Robustness analysis: Bootstrap your model coefficient estimates (use at least 500 replications). Provide a histogram of the bootstrapped estimates and comment on the findings. How do they compare to your OLS estimates?
```{r bootstrap, message=FALSE} 
library(boot)
boot_fn <- function(data, indices) {   
  d <- data[indices, ]   
  m <- lm(formula(final_model), data = d)   
  return(coef(m)) }  
set.seed(1) 
boot_out <- boot(data = df, statistic = boot_fn, R = 500)
boot_out
```
```{r boot-df} 
boot_coefs <- as.data.frame(boot_out$t) 
colnames(boot_coefs)<- names(coef(final_model)) 
head(boot_coefs)
```
#Histogram for log_income
```{r boot-hist-income} 
hist(boot_coefs$log_income, breaks = 30, main = "Bootstrap: log_income", xlab = "Estimate") 
abline(v = coef(final_model)["log_income"], lwd = 3, col = "red")
```
#Histogram for log_population 
```{r boot-hist-pop} 
hist(boot_coefs$log_population, breaks = 30, main = "Bootstrap: log_population", xlab = "Estimate") 
abline(v = coef(final_model)["log_population"], lwd = 3, col = "red")
```
#Histogram for unemployment_rate 
```{r boot-hist-unemp} 
hist(boot_coefs$unemployment_rate, breaks = 30, main = "Bootstrap: unemployment_rate", xlab = "Estimate") 
abline(v = coef(final_model)["unemployment_rate"], lwd = 3, col = "red")
```
Given the histograms, the bootsrap results indicate that income is the most stable and robust predictor of the median home vales, followed by population. Unemployment rate shows a weaker robsutness and more variability. 

### 5.6 Model validation: Use cross-validation to evaluate your model’s performance (5-fold or 10-fold) 
```{r cv-make-agec2}
library(dplyr)

df <- df %>%
mutate(
  med_age_c = med_age - mean(med_age, na.rm = TRUE),
  age_c_2  = med_age_c^2
)
```
```{r cv-final-model}
# Log dependent variable
Y <- log(df$med_home_value)

final_model <- lm(
 Y ~ log_income + log_population + unemployment_rate + vacancy_rate +
   college_share + med_age + age_c_2,
 data = df
)
```
```{r cv-clean-data}
cv_df <- df %>%
  dplyr::select(
    med_home_value, log_income, log_population, unemployment_rate,
    vacancy_rate, college_share, med_age, age_c_2
) %>%
tidyr::drop_na()
```
```{r cv-5fold}
library(caret)
set.seed(123)
train_control <- trainControl(method = "cv", number = 5)

cv_fit <- train(
 log(med_home_value) ~ log_income + log_population + unemployment_rate +
   vacancy_rate + college_share + med_age + age_c_2,
 data = cv_df,
 method = "lm",
 trControl = train_control
)

cv_fit
```
```{r cv-results}
cv_fit$results
```
We evaluated the predictive performance of the final model using the 5-fold cross validation. The RMSE is 0.245 ans the MAE is 0.192, showing relatively small prediction errors. The R^2 is 0.703, suggesting that the model explains approximately 70% of the variation in log median home values in out of sample data. Overall, these results indicate that the model generalizes well. 

### 5.7 Robust SE
```{r robust-vs-ols}
library(lmtest)
library(sandwich)

# OLS (usual) coefficient table
ols_table <- coeftest(final_model)

# Robust (HC1) coefficient table
robust_table <- coeftest(final_model, vcov = vcovHC(final_model, type = "HC1"))

ols_table
robust_table
```
We use HC1 robust standard errors to ensure valid inderence under possible heteroskedasticity. The coefficient signs and statistical significance is are pretty consistent with the OLS results. 


## 6. Results and Economic Interpretation

### Regression Table
```{r regression table}
install.packages("data.table")
install.packages("modelsummary")
library(modelsummary)

summary(model3.2)

modelsummary(model3.2)
```

### Coefficient Interpretation and Economic Significance 

Certain variables are in the log-log form which indicates that the coefficients represent the elasticities of the parameters. To illustrate, the coefficient of log_income is 1.604, meaning that a 10% change in income leads to a 16% change in median house values, holding all other variables constant. Therefore, they share a strong positive correlation as the coefficient is larger than one. This strong response means that income is a significant contributor to variation across metropolitan statistical areas. Population's coefficient is -0.021 which means that a 10% population increase would lead to a 0.21% decrease in median house values. This change is negligible compared to income meaning that population is not a driving factor in changes to housing prices. 

The other variables are in log-linear form. To demonstrate, unemployment's coefficient of 8.755 means that a 1 percentage-point increase in unemployment rate leads to an 8.76% increase in housing prices. Thus, this effect is highly significant meaning that unemployment rate be a driving force in housing prices. The coefficient of vacancy rate is 0.345 which means that a percentage increase in vacancy rate leads to an increase of 0.345%, a moderate change that is not as significant compared to unemployment. College share's coefficient of 2.552 means that a percentage point increase in college-educated residents leads to an increase of around 2.55% which indicates the strong effect between housing and housing demand from college educated residents. Median age has a coefficient of 0.004 which means that a year increase in median age increases housing values in metropolitan statistical areas is approximately 0.4% which is a smaller effect. The coefficient of the quadratic term for age indicates that the nonlinear effect of age is negligible. 

Highly statistically significant variables include log_income, unemployment rate, and college share with p-values less than 0.001 with the quadratic age term being a moderately signficant variable as its p-value is less than 0.05. This suggests that income, unemployment rate, and college share have the strongest impact on housing prices whereas population, vacancy rate, and median age are not as influential. 

### Interpretation of Confidence Intervals
```{r confidence intervals}

confint(model3.2, level = 0.95)

```
The confidence intervals for several key variables supports the statistical significance the hold. For instance, the 95% confidence interval for income elasticity spans from around 1.45 to 1.75 which shows a strong positive relationship between income and housing prices. Similarly, the 95% confidence interval for unemployment rate ranges from approximately 6.97 to 10.53 which reinforces the strong positive effect that unemployment rate holds. Finally, the college share 95% confidence interval beings at around 1.92 and 3.18 which shows significant influence of education level. All of these confidence intervals do not exhibit a sign change which leads to more confidence regarding the effect that these parameters have on housing values in MSAs. 

On the other hand, population, vacancy rate, and median age all include zero, indicating that the effect that these parameters have on housing prices is not statistically significant at this level.  

### Interpretation of R2 and Model Fit

R-squared equals 0.7159 which shows that around 71.6% of the variation of housing prices in metropolitan statistical areas is explained by the variables used in this model. This means that the model has strong explanatory power as it captures influential factors that affect housing value such as income, unemployment rate, and education level. The adjusted R-squared is at 0.7107 which suggests that the model is a strong fit even while accounting for the number of parameters included. 

### Answers to Research Questions

Housing prices in metropolitan statistical areas (MSAs) in the United States are significantly impacted by income level, unemployment rate, and educational attainment. As shown by the model, increases in income are related to higher housing prices with the largest effect as a 1% increase in income is related to a 1.60% increase in housing prices. This elasticity suggests that housing prices are highly influenced by local economic status of its residents. In like manner, unemployment rate has a significant effect as well as a 1 percentage point increase in unemployment rate leads to about an 8.76% increase in housing prices. This indicates that local labor market conditions have a strong relationship with housing market prices. Finally, the share of college educated occupants has a positive and statistically significant effect which means that populations with higher levels of education are associated with higher housing prices. All in all, stronger labor markets and greater housing demand, which are captured by these parameters, are associated with higher prices in metropolitan statistical areas. 

Nevertheless, parameters such as population size, vacancy rates, and median age do not seem to have statistically significant effects as their p-values, coefficients, and confidence intervals indicate that other variables are more influential. In effect, the model reflects that housing prices reflect the local economy as well as the level of human capital as seen by income and education attainment rather than population size and demographic characteristics of its residents.  

### Conclusions and Policy Recommendations

This model suggests that housing prices in metropolitan statistical areas in the United States are mainly affected by local socio-economic conditions. Income and educational attainment highly influence housing price variation across these areas. Due to the high elasticity of housing prices, demand for housing rises greatly in areas with higher average incomes. Furthermore, the level of education attainment has a strong relationship on housing prices which means that areas with higher levels capital reflect higher housing demand, leading to higher housing values. Another important factor is unemployment rate which suggests that labor market conditions also play a role in housing value levels.

MSAs that are economically prosperous may exhibit rapidly increasing housing demand and housing prices. This could indicate the need for policies to increase housing supply such as less restrictive zoning policies, subsidies to housing construction, or investment in affordable housing programs. Not to mention, the impact of labor market conditions also suggests that policies that lead to increased stability for labor markets and greater economic development may lead to decreased housing costs and make housing more accessible for residents. 

## 7. Limitations and Extensions
## 8. Code Quality and Reproducibility

## Sources
Blanchard, Olivier Jean & Katz, Lawrence F. 1992. "Regional Evolutions," Brookings Papers on Economic Activity, Economic Studies Program, The 
Brookings Institution, vol. 23(1), 1-76.
Diamond, Rebecca. 2016. "The Determinants and Welfare Implications of US Workers' Diverging Location Choices by Skill: 1980-2000." American Economic Review, 106 (3), 479–524.
Mian, Atif, and Amir Sufi. 2011. "House Prices, Home Equity-Based Borrowing, and the US Household Leverage Crisis." American Economic Review, 101 (5), 2132–56.
Saiz, Albert. 2010 “The Geographic Determinants of Housing Supply.” The Quarterly Journal of Economics, Volume 125, Issue 3, 1253–1296.





